Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-09-03 Manuscript/TEM_revised_manuscrip_highlithed.tex
Cleaned up manuscript folder

@@

\newpage
\section{Introduction}
Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, many large-scale macroevolutionary studies focus solely on living species (e.g. \citealp{meredithimpacts2011,jetzthe2012}).
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}.
This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}.
To do this, however, we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments \citep[e.g.][]{pyrondivergence2011,ronquista2012,BEASTmaster}.

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
These approaches differ mainly in how they treat fossil taxa and their data.
One can use fossils as tips or as nodes in the phylogeny, and can use only the age of the fossils, only the morphology of the fossils, or age and morphology jointly.
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by palaeontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
Because fossil taxa do not usually have available DNA, only fossil occurrence dates are used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the ``best'' approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}.
Neither approach, however, uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}.
This approach treats every taxa as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny \citep[known as tip-dating;][]{ronquista2012}, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}.
The Total Evidence method is becoming an increasingly popular way of adding fossil taxa to phylogenies \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014,Arcila2015131}.
Although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires both molecular and morphological data, both of which can be difficult (or impossible) to collect for every living and fossil taxon in the tree.
Morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealp{O'Leary08022013} \textit{vs.} \citealp{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record.
For example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts \citep{sansomfossilization2013}; and molecular data are (nearly) always unavailable.
Therefore Total Evidence matrices are likely to contain a large proportion of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

Although missing data do not appear be a major problem in molecular and morphological matrices separately \citep[as long as enough data overlap in each case, and missing data are not phylogenetically biased;][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework \citep[i.e. using both molecular and morphological data;][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.
These previous studies assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data \citep{Wiens01102005}; or (2) removing both molecular and some morphological data from living taxa to create artificial fossils \citep{manosphylogeny2007,pattinsonphylogeny2014}.
Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
The way these studies were conducted, however, means that their conclusions are not generally applicable across all scenarios involving missing data in Total Evidence phylogenies.
For example, using an empirical (rather than simulation based) approach limits their conclusions to studies with similar distributions of data across species in the phylogeny.
Additionally, one of the three previous studies did not include fossil taxa in their analyses, so their results cannot be used to make conclusions about how missing data may influence the placement of fossils \citep{wiensmissing2003}. The other two studies did include fossil taxa, but used the patchiness of the fossil record to determine how to remove data from their matrices \citep{manosphylogeny2007,pattinsonphylogeny2014}. Data for living species are unlikely to be missing in this patchy way, instead full molecular data with the complete absence of morphological data is a likely pattern \citep{GuillermeCooperMissing}.
Finally, these previous studies mainly focused on how missing data in fossil taxa affect the placement of fossils, ignoring the effects of missing data in living species \citep{manosphylogeny2007,pattinsonphylogeny2014}.

In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living taxa with morphological data, fossil data, all data for certain characters and the combination of these three aspects.
This is an advance on previous studies because we use large-scale simulations and analyse the effects of three distinct aspects of missing data thus focusing on both neontological and palaeontological parts of the matrix.
In addition, we test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian phylogenies: (i) the conservation of clades \citep[based on the Robinson-Foulds distance;][]{RF1981} and (ii) the displacement of wild-card taxa \citep[based on the Triplets distance;][]{critchlowthe1996} rather than just a single measure of clade conservation or clade support \citep[cf.][]{Wiens01102005,pattinsonphylogeny2014}.

We focus on the effects of missing data on our ability to recover tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}.
Although branch length estimation is also important \citep[namely for timing extinction and/or speciation events; e.g.][]{ronquista2012}, we do not consider branch lengths in this study.
This is partially due to difficulties with simulating branch lengths and topology simultaneously, but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\section{Materials and Methods}
To explore how missing data in the morphological partition of Total Evidence matrices influences tree topology, we used the following protocol (Fig. 1):
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true'' tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete'' matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete'' matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the number of morphological characters ($N_{C}$). We call the resulting 125 matrices ``missing-data'' matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete'' matrix and from the 125 ``missing-data'' matrices resulting in one tree generated from a matrix with no missing data (hereafter called the ``best'' tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data'' trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best'' tree to the ``missing-data'' trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $N_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

@@

Next, we generated a molecular and a morphological matrix from the ``true'' tree.
The molecular matrix was simulated from the ``true'' tree using the R package phyclust v. 0.1-14 \citep{chen2011}.
The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}.
The substitution rates were selected from a gamma distribution with an ($\alpha$) shape of 0.5 \citep{yangamong-site1996}.
In practice, a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal \citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this $\alpha$ value to be consistent with our protocol for simulating morphological characters (see below).
This model and these parameter settings strike a balance between realism for empirical datasets \citep[e.g.][]{douadycomparison2003,kellymolecular2014} and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years).
All the molecular information for fossil taxa was replaced by missing data ("?").

We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa.
We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}.
This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
We used low evolutionary rate parameters to be consistent with the molecular rate parameters, to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wrightbayesian2014}.
Topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model \citep{lewisa2001,wrightbayesian2014}.
Note, however, that \cite{wrightbayesian2014} have shown that low morphological rates ($<$ 0.5) increase variance in topological error, but we discarded simulations with such topological error by selecting only matrices with a ``fair'' phylogenetic signal \citep[see Estimating phylogenies section below;][]{zanderminimal2004} so this should not influence our results.

@@

\subsection{Removing data}
To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:
\begin{enumerate}
\item{The proportion of missing living taxa ($M_L$).}
This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species are rarely collected \citep{GuillermeCooperMissing}.
Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
\item{The proportion of missing data in the fossil record ($M_F$).}
This missing data parameter represents the completeness of the fossil record.
Due to preservation biases, missing data for fossil taxa are common \citep{sansomfossilization2013}.
In practice, we randomly removed a proportion of data from across the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose most of their data at this parameter value (down to a minimum threshold of 5\% available data; see below).
\item{The number of morphological characters for both living and fossil taxa ($N_C$).}
This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large.
Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed entire characters from the morphological matrix reducing it to: 100, 90, 75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).
\end{enumerate}

Each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living data; $M_F$ removes cells from the fossil data; and $N_C$ removes columns across both living and fossil data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data'' matrices.
Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxa.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).

@@

\subsection{Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best'' tree inferred from the ``complete'' matrix and the ``missing-data'' trees inferred from the 125 matrices with various amounts of missing data.
The ``true'' tree was used to generate the ``complete'' matrix and reflects the ``true'' evolutionary history in our simulations.
The ``best'' tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods.
In real world situations, the ``true'' tree is never available to us because we cannot know the true evolutionary history of a clade \citep[except in very rare circumstances, e.g.][]{rozen2005}.
We compare ``best'' trees to ``missing data'' trees but could also compare ``true'' trees to the ``missing data'' trees.
In practice, the difference between the ``best'' trees and the ``missing data'' trees represents the effect of our missing data parameters and of the phylogenetic methods used to infer the ``missing data'' trees.
The difference between the ``true'' and the ``missing data'' trees, however, represents the effect of our parameters used to generate the ``true'' tree and the algorithms used to generate the ``complete'' matrix as well as the effect of our missing data parameters and the phylogenetic methods used.
Because the main aim of this study is to look at the effect our missing data parameters on topological recovery, we chose to represent only the comparisons between the ``best'' trees and ``missing data'' trees.
The results of the comparisons of the ``true'' tree and the ``missing data'' trees are available in Appendix B.
Note that this makes little difference to our overall results.

@@

\subsubsection{Missing data for fossil taxa ($M_{F}$)}
When the overall proportion of data for the fossil taxa ($M_{F}$) decreases, this also reduces the probability of morphological characters for fossil taxa overlapping with the ones for living taxa.
This can lead to difficulties for the placement of certain taxa in the tree.
It is important, however, to note that even though the number of displaced wildcard taxa increases (i.e. decrease of Normalised Triplets metric) with increasing missing data in this parameter, clade conservation (i.e. Normalised Robinson-Foulds metric) is still relatively good (mode = 0.72) when the proportion of missing data are high ($M_{F}$ = 75\%).
These results are in agreement with \cite{manosphylogeny2007} where as few as 16 characters were sufficient for correctly assigning artificial fossils to their correct clade.

>>>>

%---------------------------------------------
%
%       THIS IS THE HIGHLIGHTED VERSION.
%       For the final version, carefully remove:
%       '\protect'
%       '\hl{' (don't forget the '}'
%
%---------------------------------------------

@@

\newpage
\section{Introduction}
Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, \hl{many large-scale macroevolutionary studies focus solely on living species} (e.g. \citealp{meredithimpacts2011,jetzthe2012}).
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}.
This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}.
To do this, however, we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments \citep[e.g.][]{pyrondivergence2011,ronquista2012,BEASTmaster}.

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies.
\hl{These approaches differ mainly in how they treat fossil taxa and their data.
One can use fossils as tips or as nodes in the phylogeny, and can use only the age of the fossils, only the morphology of the fossils, or age and morphology jointly.}
Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{Hennig1966,felsenstein2004}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty.
Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species.
\hl{Because fossil taxa do not usually have available DNA, only fossil occurrence dates are used to time calibrate phylogenies }\citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the ``best'' approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}.
Neither approach, however, uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}.
This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny \citep[known as tip-dating;][]{ronquista2012}, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}.
\hl{The Total Evidence method is becoming an increasingly popular way of adding fossil taxa to phylogenies} \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014,Arcila2015131}.
Although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires both molecular and morphological data, both of which can be difficult (or impossible) to collect for every living and fossil taxon in the tree.
Morphological data for living taxa are rarely collected when molecular data are available (e.g. \citealp{O'Leary08022013} \textit{vs.} \citealp{meredithimpacts2011}), and for fossil taxa, data can only be collected from features preserved in the fossil record.
For example, in vertebrates, the hardest parts of the skeleton are more often preserved than soft parts \citep{sansomfossilization2013}; and molecular data \hl{are} (nearly) always unavailable.
Therefore \hl{Total Evidence matrices are likely to contain a large proportions of missing data} that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

Although missing data do not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework }\citep[\hl{i.e. using both molecular and morphological data;}][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.
\hl{These previous studies assessed the effect of missing data on topology by either (1) comparing a dataset with missing data to subsets without missing data }\citep{Wiens01102005}\hl{; or (2) removing both molecular and some morphological data from living taxa to create artificial fossils }\citep{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Both approaches have shown that missing data are not a major problem and should not be an obstacle to combining both living and fossil species in the same phylogenies.
The way these studies were conducted, however, means that their conclusions are not generally applicable across all scenarios involving missing data in Total Evidence phylogenies.
For example, using an empirical (rather than simulation based) approach limits their conclusions to studies with similar distributions of data across species in the phylogeny.
Additionally, }\cite{Wiens01102005}\hl{ did not include fossil taxa in their analyses, so their results cannot be used to make conclusions about how missing data may influence the placement of fossils.}
\cite{manosphylogeny2007,pattinsonphylogeny2014}\hl{ include fossil taxa, but used the patchiness of the fossil record to determine how to remove data from their matrices. 
Data for living species are unlikely to be missing in this patchy way, instead full molecular data with the complete absence of morphological data is a likely pattern}\citep{GuillermeCooperMissing}.
\hl{Finally, }\cite{manosphylogeny2007}\hl{ and }\cite{pattinsonphylogeny2014}\hl{ mainly focused on how missing data in fossil taxa affect the placement of fossils, ignoring the effects of missing data in living species.}

\hl{In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by removing living taxa with morphological data, fossil data, all data for certain characters and the combination of these three aspects.
This is an advance on previous studies because we use large-scale simulations and analyse the effects of three distinct aspects of missing data thus focusing on both neontological and paleontological parts of the matrix.
In addition, we test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian phylogenies: (i) the conservation of clades }\citep[\hl{based on the Robinson-Foulds distance;}][]{RF1981}\hl{ and (ii) the displacement of wild-card taxa }\citep[\hl{based on the Triplets distance;}][]{critchlowthe1996}\hl{ rather than just a single measure of clade conservation or clade support }\citep[\hl{cf.}][]{Wiens01102005,pattinsonphylogeny2014}.

We focus on the effects of missing data on our ability to recover tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}.
Although branch length estimation is also important \citep[namely for timing extinction and/or speciation events; e.g.][]{ronquista2012}, we do not consider branch lengths in this study.
This is partially due to difficulties with simulating branch lengths and topology simultaneously, \hl{but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints }\citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\section{Materials and Methods}
To explore how missing data in \hl{the morphological partition of Total Evidence matrices} influences tree topology, \hl{we used the following protocol (Fig. 1):}
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true'' tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete'' matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete'' matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the number of morphological characters ($N_{C}$). We call the resulting 125 matrices ``missing-data'' matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete'' matrix and from the 125 ``missing-data'' matrices resulting in one tree generated from a matrix with no missing data (hereafter called the ``best'' tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data'' trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best'' tree to the ``missing-data'' trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $N_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

@@

Next, we generated a molecular and a morphological matrix from the ``true'' tree.
The molecular matrix was \hl{simulated} from the ``true'' tree using the R package phyclust v. 0.1-14 \citep{chen2011}.
The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \citep{douadycomparison2003}.
\hl{The substitution rates were selected from a gamma distribution with an ($\alpha$) shape of 0.5} \citep{yangamong-site1996}.
\hl{In practice, a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal }\citep{Hassanin1998611,EstoupHomoplasy}.
\hl{Also, we chose this $\alpha$ value to be consistent with our protocol for simulating morphological characters (see below).}
\hl{This model and these parameter settings strike a balance between realism for empirical datasets }\citep[\hl{e.g.}][]{douadycomparison2003,kellymolecular2014}\hl{ and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years).}
All the molecular information for fossil taxa was replaced by missing data ("?").

\hl{We simulated the morphological matrix using the rTraitDisc function from the R package ape v. 3.0-11} \citep{paradisape:2004}\hl{ to generate a matrix of 100 character sites for 51 taxa.}
\hl{We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within).}
We then ran an independent discrete character simulation for each character using the ``true'' tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}.
This method allows us to have only two parameters for each character: the number of states and the evolutionary rate.
For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5.
\hl{We used low evolutionary rate parameters to be consistent with the molecular rate parameters, to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal }\citep{wrightbayesian2014}.
\hl{Topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model }\citep{lewisa2001,wrightbayesian2014}.
\hl{Note, however, that } \cite{wrightbayesian2014}\hl{ have shown that low morphological rates ($<$ 0.5) increase variance in topological error, but we discarded simulations with such topological error by selecting only matrices with a ``fair'' phylogenetic signal }\citep[\hl{see Estimating phylogenies section below;}][]{zanderminimal2004}\hl{ so this should not influence our results.}

@@

\subsection{Removing data}
\hl{To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:}
\begin{enumerate}
\item{\hl{The proportion of missing living taxa ($M_L$).}}
\hl{This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species are rarely collected }\citep{GuillermeCooperMissing}.
\hl{Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.}
\item{\hl{The proportion of missing data in the fossil record ($M_F$).}}
\hl{This missing data parameter represents the completeness of the fossil record.
Due to preservation biases, missing data for fossil taxa are common }\citep{sansomfossilization2013}.
\hl{In practice, we randomly removed a proportion of data from across the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose most of their data at this parameter value (down to a minimum threshold of 5\% available data; see below).}
\item{\hl{The number of morphological characters for both living and fossil taxa ($N_C$).}}
\hl{This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large.
Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed entire characters from the morphological matrix reducing it to: 100, 90, 75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).}
\end{enumerate}

\hl{Each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living data; $M_F$ removes cells from the fossil data; and $N_C$ removes columns across both living and fossil data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.}

\hl{We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data'' matrices.
Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).}

@@

\subsection{Estimating phylogenies}
From the resulting matrices we generated two types of trees: the ``best'' tree inferred from the ``complete'' matrix and the ``missing-data'' trees inferred from the 125 matrices with various amounts of missing data.
The ``true'' tree was used to generate the ``complete'' matrix and reflects the ``true'' evolutionary history in our simulations.
The ``best'' tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods.
In real world situations, the ``true'' tree is never available to us because we cannot know the true evolutionary history of a clade \citep[except in very rare circumstances, e.g.][]{rozen2005}.
\hl{We compare ``best'' trees to ``missing data'' trees but could also compare ``true'' trees to the ``missing data'' trees.
In practice, the difference between the ``best'' trees and the ``missing data'' trees represents the effect of our missing data parameters and of the phylogenetic methods used to infer the ``missing data'' trees.
The difference between the ``true'' and the ``missing data'' trees, however, represents the effect of our parameters used to generate the ``true'' tree and the algorithms used to generate the ``complete'' matrix as well as the effect of our missing data parameters and the phylogenetic methods used.
Because the main aim of this study is to look at the effect our missing data parameters on topological recovery, we chose to represent only the comparisons between the ``best'' trees and ``missing data'' trees.
The results of the comparisons of the ``true'' tree and the ``missing data'' trees are available in Appendix B.
Note that this makes little difference to our overall results.}

@@

\subsubsection{Missing data for fossil taxa ($M_{F}$)}
When the overall proportion of data for the fossil taxa ($M_{F}$) decreases, this also reduces the probability of morphological characters for fossil taxa overlapping with the ones for living taxa.
This can lead to difficulties for the placement of certain taxa in the tree.
It is important, however, to note that even though the number of displaced wildcard taxa increases (i.e. decrease of Normalised Triplets metric) with increasing missing data in this parameter, clade conservation (i.e. Normalised Robinson-Foulds metric) is still relatively good (mode = 0.72) when the proportion of missing data are high ($M_{F}$ = 75\%).
\hl{These results are in agreement with }\cite{manosphylogeny2007}\hl{ where as few as 16 characters were sufficient for correctly assigning artificial fossils to their correct clade.}
